Sparse Function-space Representation \\ of Neural Networks

### 2023-05-17 paper/main.tex
updaetd rows of table

%Current state of affairs 
Recent techniques (\eg, \cite{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable}) apply a Laplace-GGN approximation\todo{Is Laplace-GGN a well-known acronym?} to convert trained neural networks into Bayesian neural networks, that can provide uncertainty without sacrificing additional resources to training \cite{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \cite{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing the behaviour \cite{cho2009kernel,meronen2020stationary}, performing probabilistic inference \cite{khan2019approximate}, and quantifying uncertainty in neural networks \cite{foong2019between}. These methods rely on the linearization of the neural network and the resultant neural tangent kernel (NTK, \cite{jacot2018neural}). The neural network is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates. They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.  
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits \todo{is 'emits' a clear enough expression here?}`dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data to avoid the computational complexity of constructing the GP over the full data set\todo{What kind of subsets? Probably should specify NTK + GP subset to be clear} \cite{immer2021scalable}, this parameterization allows consolidating contributions from {\em all} training data into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP now predicts in the same space as the original trained neural network \todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach to condition on new incoming data into our model crucially without need of retraning it(see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

%Need uncertainty and adaptive methods
%Dual formulation in GPs space solves this
%Talk about planning and exploration in RL.

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, that linearize a trained neural network around MAP weights and approximates the neural network's Hessian using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). 
%\todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}
%The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation . 
While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. \todo{Reads a bit repetitive to start two sentences in a row by This, could maybe be combined in one.} This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains \todo{Maybe unexplored instead of unclear?} unclear without resorting to subset methods \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\begin{table}[t!] 
  \centering\scriptsize
  \caption{Metrics for supervised learning with image data. SVGP is our method, and the number inside parentheses is the number of inducing points. SVGP NN is a modification of our method, where the mean is from the NN directly and the variance is from the GP model.} 
	\label{tbl:imagesuper}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{6pt}
	\setlength{\tblw}{0.095\textwidth}

@@

    % THE TABLE CAN BE FORMATTED MORE LIKE THIS
    % \begin{tabular}{l l C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    % \toprule
    % & Method & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & OOD-AUC~$\uparrow$  \\
    % \midrule
    % \multirow{2}{*}{FMNIST}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \midrule
    % \multirow{2}{*}{CIFAR-10}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \bottomrule
    % \end{tabular}
  \newcommand{\myline}{\protect\tikz[baseline=-.5ex,line width=1pt]\protect\draw[draw=black!50](0,0)--(7em,0);}
  \begin{tabular}{l c C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    \toprule
            &     &  \multicolumn{3}{c}{\myline~MNIST~\myline} & \multicolumn{3}{c}{\myline~FMNIST~\myline} \\
     Method & $M$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ \\
    \midrule
    %\multirow{2}{*}{MNIST}
    MAP & & \val{98.22}{0.13} & \val{0.061}{0.004} & \val{0.006}{0.001}
          & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    BNN predictive \cite{immer2021improving} & & \val{93.14}{0.05} & \val{0.304}{0.002} & \val{0.111}{0.003}
                                               & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    BNN predictive \cite{ritter2018kfac} & & \val{93.03}{0.13} & \val{0.369}{0.003} & \val{0.168}{0.001}
                                           & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    GLM predictive \cite{immer2021improving} & & \val{98.40}{0.05} & \val{0.054}{0.002} & \val{0.007}{0.001}
                                               & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    GP predictive \cite{immer2021improving} & $3200$ & \val{98.22}{0.13} & \val{0.058}{0.001} & \val{0.003}{0.000}
                                                     & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\

    GP predictive NN & $1024$ &  \val{97.70}{0.07} & \val{0.080}{0.002} & \val{0.011}{0.001}
                              &  \val{0}{0.07} & \val{0}{0.002} & \val{0}{0.001} \\

    GP predictive & $1024$  & \val{76.28}{1.85} & \val{0.702}{0.029} & \val{0.070}{0.022}
                            & \val{0}{1.85} & \val{0}{0.029} & \val{0}{0.022} \\
    \our NN & $1024$ &  \val{98.02}{0.12} & \val{0.066}{0.003} & \val{0.007}{0.001}
                     &  \val{0}{0.0} & \val{0}{0.003} & \val{0}{0.001} \\
    \our & $1024$ & \val{98.07}{0.06} & \val{0.063}{0.003} & \val{0.005}{0.001}
                  & \val{0}{0.06} & \val{0}{0.003} & \val{0}{0.001} \\
%    \multirow{2}{*}{FMNIST}
%    & MAP & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    %& BNN predictive \cite{immer2021improving} & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    % & BNN predictive \cite{ritter2018kfac} & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    % & GLM predictive \cite{immer2021improving} & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    %& GP predictive \cite{immer2021improving} $m=3200$ & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\
    % & GP predictive NN $m=1024$ &  \val{0}{0.07} & \val{0}{0.002} & \val{0}{0.001} \\
    % & GP predictive $m=1024$  &  \val{0}{1.85} & \val{0}{0.029} & \val{0}{0.022} \\
    % & \our NN $m=1024$ &  \val{0}{0.0} & \val{0}{0.003} & \val{0}{0.001} \\
    % & \our $m=1024$ &  \val{0}{0.06} & \val{0}{0.003} & \val{0}{0.001} \\
    \bottomrule
  \end{tabular}

>>>>

%Current state of affairs 
Recent techniques (\eg, \cite{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable}) apply the Laplace-GGN approximation\todo{Is Laplace-GGN a well-known acronym?} to convert a trained neural networks into a Bayesian neural network, that can provide uncertainty without sacrificing additional resources to training \cite{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \cite{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing the behaviour \cite{cho2009kernel,meronen2020stationary}, performing probabilistic inference \cite{khan2019approximate}, and quantifying uncertainty in neural networks \cite{foong2019between}. These methods rely on the linearization of the neural network and the resultant neural tangent kernel (NTK, \cite{jacot2018neural}). The neural network is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates. They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.  
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits \todo{is 'emits' a clear enough expression here?}`dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets \todo{What kind of subsets? Probably should specify NTK + GP subset to be clear} \cite{immer2021scalable}, this parameterization allows consolidating contributions from {\em all} training data into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP now predicts in the same space as the original trained neural network \todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as \our---a sparse GP derived from a trained neural network \todo{I assume FR is for functional representation? Could be written here for extra clarity}. As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.\looseness-2
%Need uncertainty and adaptive methods
%Dual formulation in GPs space solves this
%Talk about planning and exploration in RL.

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, linearizing a trained neural network around optimal weights. \todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. \todo{Reads a bit repetitive to start two sentences in a row by This, could maybe be combined in one.} This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains \todo{Maybe unexplored instead of unclear?} unclear without resorting to subset methods \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\begin{table}[t!] 
  \centering\scriptsize
  \caption{Metrics for supervised learning with image data. SVGP is our method, and the number inside parentheses is the number of inducing points. SVGP NN is a modification of our method, where the mean is from the NN directly and the variance is from the GP model.} 
	\label{tbl:imagesuper}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{6pt}
	\setlength{\tblw}{0.15\textwidth}

@@

    % THE TABLE CAN BE FORMATTED MORE LIKE THIS
    % \begin{tabular}{l l C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    % \toprule
    % & Method & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & OOD-AUC~$\uparrow$  \\
    % \midrule
    % \multirow{2}{*}{FMNIST}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \midrule
    % \multirow{2}{*}{CIFAR-10}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \bottomrule
    % \end{tabular}
  \begin{tabular}{l c C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    \toprule
     Method & M &  & MNIST &  &  & FMNIST &  \\
            &   & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ \\
    \midrule
    %\multirow{2}{*}{MNIST}
    MAP & & \val{98.22}{0.13} & \val{0.061}{0.004} & \val{0.006}{0.001}
          & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    BNN predictive \cite{immer2021improving} & & \val{93.14}{0.05} & \val{0.304}{0.002} & \val{0.111}{0.003}
                                               & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    BNN predictive \cite{ritter2018kfac} & & \val{93.03}{0.13} & \val{0.369}{0.003} & \val{0.168}{0.001}
                                           & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    GLM predictive \cite{immer2021improving} & & \val{98.40}{0.05} & \val{0.054}{0.002} & \val{0.007}{0.001}
                                               & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    GP predictive \cite{immer2021improving} & $3200$ & \val{98.22}{0.13} & \val{0.058}{0.001} & \val{0.003}{0.000}
                                                     & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\

    GP predictive NN & $1024$ & \val{97.70}{0.07} & \val{0.080}{0.002} & \val{0.011}{0.001}
                              & \val{91.47}{0.41} & \val{0.289}{0.009} & \val{0.033}{0.003} \\

    GP predictive & $1024$  & \val{76.28}{1.85} & \val{0.702}{0.029} & \val{0.070}{0.022}
                            & \val{83.98}{0.35} & \val{0.503}{0.012} & \val{0.052}{0.006} \\
    \our NN & $1024$ &  \val{98.02}{0.12} & \val{0.066}{0.003} & \val{0.007}{0.001}
                     & \val{91.93}{0.45} & \val{0.267}{0.011} & \val{0.028}{0.003} \\
    \our & $1024$ & \val{98.07}{0.06} & \val{0.063}{0.003} & \val{0.005}{0.001}
                  & \val{91.56}{0.35} & \val{0.253}{0.006} & \val{0.012}{0.001} \\
    \midrule
%    \multirow{2}{*}{FMNIST}
%    & MAP & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    %& BNN predictive \cite{immer2021improving} & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    % & BNN predictive \cite{ritter2018kfac} & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    % & GLM predictive \cite{immer2021improving} & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    %& GP predictive \cite{immer2021improving} $m=3200$ & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\
    % & GP predictive NN $m=1024$ &  \val{0}{0.07} & \val{0}{0.002} & \val{0}{0.001} \\
    % & GP predictive $m=1024$  &  \val{0}{1.85} & \val{0}{0.029} & \val{0}{0.022} \\
    % & \our NN $m=1024$ &  \val{0}{0.0} & \val{0}{0.003} & \val{0}{0.001} \\
    % & \our $m=1024$ &  \val{0}{0.06} & \val{0}{0.003} & \val{0}{0.001} \\
    \bottomrule
  \end{tabular}
